---
title: "STAT 1341 Final"
author: "Ian Keller"
date: "2023-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading in NHL Data
```{r}
data <- read.table("nhl_data.csv", header = TRUE, sep =",")
advanced_stats <- read.table("2017_team_advanced_stats.csv", header = TRUE, sep = ",")
```

Filtering Data to 2016-2017 NHL Season
```{r}
library(dplyr)
nhl_2017 <- data %>%
  filter(season == 2016)
```
Visualizing relevant statistics
```{r}
library(ggplot2)
ggplot(advanced_stats, aes(x = reorder(Team, xGF.), y = xGF., fill = xGF.)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "blue", high = "red") + # Colors can be adjusted
  labs(x = "Team", y = "xGF%", title = "Expected Goals For Percentage (xGF%)") +
  theme_minimal()

```
Visualizing PDO and Team
```{r}
ggplot(advanced_stats, aes(x = reorder(Team, PDO), y = PDO, group = 1)) +
  geom_line() +
  geom_point() +
  coord_flip() +
  labs(x = "Team", y = "PDO", title = "PDO by Team") +
  theme_minimal()
```
```{r}
ggplot(advanced_stats, aes(x = reorder(Team, PDO), y = PDO)) +
  geom_point() +
  coord_flip() +
  labs(x = "Team", y = "PDO", title = "PDO Distribution Across Teams") +
  theme_minimal()
```
Bar Chart
```{r}
ggplot(advanced_stats, aes(x = reorder(Team, PDO), y = PDO, fill = PDO)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "blue", high = "red") + # Gradient color based on PDO values
  coord_flip() + # Flip the coordinates for better readability of team names
  labs(x = "Team", y = "PDO", title = "PDO by Team") +
  theme_minimal()
```
xGF vs GF for each team
```{r}
library(tidyverse)
df_long <- advanced_stats %>%
  select(Team, xGF, GF) %>%
  pivot_longer(cols = c(xGF, GF), names_to = "Metric", values_to = "Value")

# Create the grouped bar chart
ggplot(df_long, aes(x = Team, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip() +
  labs(x = "Team", y = "Value", title = "Comparison of xGF and GF for Each Team") +
  scale_fill_brewer(palette = "Set1") + # Use a distinct color palette
  theme_minimal()
```
1
```{r}
advanced_stats$GF_diff = advanced_stats$GF - advanced_stats$xGF

# Create the bar chart
ggplot(advanced_stats, aes(x = reorder(Team, GF_diff), y = GF_diff, fill = GF_diff)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) + # Color gradient
  labs(x = "Team", y = "GF - xGF", title = "Difference Between GF and xGF for Each Team") +
  theme_minimal()
```
Five Number Summary of xGF
```{r, dpi=300}
hist(advanced_stats$xGF., breaks =6, prob = TRUE, 
     main = "Histogram of Expected Goals for Percentage", xlab = "Expect Goals For Percentage")
lines(density(advanced_stats$xGF.), col =1, lwd =2)
```
Calculating Mean, Sd, and Median of Variables of interest
```{r}
paste("Mean of xGF%", mean(advanced_stats$xGF.))
paste("Sd of xGF%", sd(advanced_stats$xGF.))
paste("Median of xGF%", median(advanced_stats$xGF.))
```
Finding Outliers
```{r}
# Calculate quartiles
Q1 <- quantile(advanced_stats$xGF., 0.25)
Q3 <- quantile(advanced_stats$xGF., 0.75)

# Compute the IQR
IQR <- Q3 - Q1

# Calculate outlier thresholds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
outliers <- subset(advanced_stats, advanced_stats$xGF. < lower_bound | advanced_stats$xGF. > upper_bound)

# Print outliers
print(outliers)

```
SCoring Distribution for the season
```{r}
nhl = read.table("nhl_data.csv", header = TRUE, sep = ",")
nhl = nhl[which(nhl$season == 2016 & nhl$home_score > -1 & nhl$game_type == "r"), ]

# Histogram of all scores with shootout goals included
nhl_scores = c(nhl$home_score, nhl$away_score)
barplot(table(nhl_scores), xlab = "NHL Goals Scored with Shootouts", ylab = "Frequency")


# Change final score of NHL overtime games to regulation score
for (i in 1:nrow(nhl)) {
    if (nhl$tie[i] == 1) {
        if(nhl$home_score[i] > nhl$away_score[i]) {
            nhl$home_score[i] = nhl$home_score[i] - 1
        } else if (nhl$home_score[i] < nhl$away_score[i]) {
            nhl$away_score[i] = nhl$away_score[i] - 1
        }
    }
}

# Histogram of all scores with shootout goals removed
nhl_scores = c(nhl$home_score, nhl$away_score)
barplot(table(nhl_scores), xlab = "NHL Goals Scored without Shootouts", ylab = "Frequency")
```
Home vs Away Scores
```{r}
library(moments)
nhl_scores = c(nhl$home_score, nhl$away_score)


# Choose from nhl_scores, nhl$home_score, or nhl$away_score

variable = nhl$away_score


# Initialize table that will hold actual number of times each number of goals was scored and the theoretical number of times teams should have scored each number of goals assuming a Poisson distribution.
counts = matrix(nrow = 2, ncol = max(variable) + 1, byrow=TRUE)  
colnames(counts) = 0:max(variable)
rownames(counts) = c('actual', 'expected')


# Tables comparing number of instances of each number of goals was scored against the expected count assuming a Poisson distribution with above mean
counts[1,1:(max(variable) + 1)] = table(variable)
proportions = counts / length(variable)


# Calculate proportion (probability) of goals according to Poisson distribution
for(k in 0:max(variable)) {
    proportions[2,k+1] = (mean(variable)^k) * exp(-mean(variable)) / factorial(k)
}


# Calculate theoretical number of instances of each number of goals according to Poisson distribution
counts[2,1:(max(variable) + 1)] = proportions[2,1:(max(variable) + 1)] * length(variable)


# Create bar graphs of goals scored and goals scored vs. theoretical number of instances of each number of goals scored according to a Poisson distribution.
barplot(counts[1,], xlab = "Away Goals", ylab = "Frequency")
barplot(proportions, xlab = "Away Goals", ylab = "Probability", ylim = c(0, 0.25), beside = TRUE)


# Calculate chi-squared components
components = (counts[1,]-counts[2,])^2/counts[2,]


# Sum chi-squared components for only goals where expected count is at least 10.  In this case, stop adding after 8 goals.  The last index will have to be manually adjusted based on the components above.
X2 = sum(components[1:9])
p = 1-pchisq(X2, 8)
```
Margin
```{r}
barplot(table(nhl$home_score - nhl$away_score), xlab = "Home Goals - Away Goals")
plot(nhl$home_score, nhl$away_score, xlab = "Home Goals", ylab = "Away Goals")
cor(nhl$home_score, nhl$away_score)
```
Home and Away Stats
```{r}
home_mean <- mean(nhl$home_score)
home_sd <- sd(nhl$home_score)

away_mean <- mean(nhl$away_score)
away_sd <- sd(nhl$away_score)

paste("home mean: ", home_mean)
paste("home sd: ", home_sd)
paste("away mean:", away_mean)
paste("away sd:", away_sd)
```

# Calculating Pythagorean Win Percentage
Creating new df
```{r}
nhl_14 <- read.table("2013-14_nhl_season.csv", header = TRUE, sep = ",")
nhl_14$season = 2013

nhl_15 <- read.table("2014-15_nhl_season.csv", header = TRUE, sep = ",")
nhl_15$season = 2014

nhl_16 <- read.table("2015-16_nhl_season.csv", header = TRUE, sep = ",")
nhl_16$season = 2015

nhl_17 <- read.table("2016-17_nhl_season.csv", header = TRUE, sep = ",")
nhl_17$season = 2016
combined_nhl <- rbind(nhl_14, nhl_15, nhl_16, nhl_17)
combined_nhl$league = "NHL"

library(dplyr)
scores <- combined_nhl %>%
  select(league, season, Team, Point.., GF, GA)

scores <- scores %>%
  rename(win_p = Point..,
         scored = GF,
         allowed = GA)

```




```{r}

# Specify league and desired seasons to find Pythagorean exponent
league = "NHL"
first_season = 2013
last_season = 2016
s = scores[which(scores$league == league & scores$season >= first_season & scores$season <= last_season & scores$win_p > 0), ]

# Fit model to obtain optimal Pythagorean exponent
model = lm(log(win_p) ~ 0 + log(scored/allowed), data = s)
summary(model)

plot(log(win_p) ~ 0 + log(scored/allowed), data = s, pch = 19, xlab = "Log(Goals for / Goals allowed)", ylab = "Log(Win Percentage)")


# Calculate actual winning percentage, Pythagorean winning percentage, and error
s$wpct = s$win_p
s$pyth_wpct = s$scored^coef(model) / (s$scored^coef(model) + s$allowed^coef(model))
s$error1 = s$wpct - s$pyth_wpct

# Plot actual winning percentage against Pythagorean winning percentage
plot(s$pyth_wpct, s$wpct, xlab = "Pythagorean Winning Percentage", ylab = "Actual Winning Percentage", pch = 19)
cor(s$pyth_wpct, s$wpct)

# Analyzing errors
hist(s$error, xlab = "Actual Win Pct. - Pythagorean Win Pct.", breaks = 15)

#Disregarding Close Game numbers
cor(s$wpct, s$pyth_wpct)

```
S for 2016
```{r}
library(dplyr)
s_2016 <- s %>%
  filter(season == 2016)
```

# Poisson ratings

```{r}
scores <- nhl_2017 %>%
  select(season, home_team, away_team, home_score, away_score, location)
```

Running the model
```{r}
# Identify unique names of all teams
teams = sort(unique(scores$home_team))
off_teams = teams
def_teams = off_teams

# Create empty data frame for margin of victory, game location, and one column for each team
data_matrix = as.data.frame(matrix(nrow = 2*nrow(scores), ncol = 2+2*length(teams)))

for (i in 1:length(off_teams)) {
    off_teams[i] = paste(off_teams[i], "Off.")
}

for (i in 1:length(off_teams)) {
    def_teams[i] = paste(def_teams[i], "Def.")
}

# Replace generic column names in 'nfl_data' data frame with columns for margin of victory, location, and the 32 team names
colnames(data_matrix) = c("Score", "Location", off_teams, def_teams)

# Replace all entries in data frame with 0
data_matrix[is.na(data_matrix)] = 0

# The following loop iterates through all 32 teams.  For each team, the first line sets the value equal to 1 if that team was at home and 0 otherwise.  The second line sets the value equal to -1 if that team was away and 0 otherwise.

for (i in 1:nrow(scores)) {
    for (j in 1:length(teams)) {
        if (scores$home_team[i] == teams[j]) {
            data_matrix[i, j+2] = 1
            data_matrix[i+nrow(scores), j+2+length(teams)] = -1
            break
        }
    }
    
    for (j in 1:length(teams)) {
        if (scores$away_team[i] == teams[j]) {
            data_matrix[i, j+2+length(teams)] = -1
            data_matrix[i+nrow(scores), j+2] = 1
            break
        }
    }
}

# Set margin of victory to be home score minus away score
data_matrix$Score = c(scores$home_score, scores$away_score)

# Set game location to be 1 if the home team was in their home stadium; set equal to 0 if the game was played at a neutral location with a designated home team.  This normally needs to be done manually in practice, but will always be provided for you in the original dataset for the purposes of this course.
data_matrix$Location = c(scores$location, 0*scores$location)
```
Running the model frfr
```{r}
# Fit Poisson regression model
model = glm(Score ~ ., data = data_matrix, family = "poisson")
summary(model)
```
This chunk extracts the team offensive and defensive coefficients and sets the baseline category equal to 0.

```{r}
coefficients = coef(model, complete = TRUE)   # Extract only coefficients from model summary

# Change baseline team's offensive and defensive ratings to 0
for (i in 1:length(coefficients)) {
    if (is.na(coefficients[i])) {
        coefficients[i] = 0
    }
}
```

This code calculates the expected goals scored in a match for each team as well as the joint probability distribution for the final scoreline.  Each team's rating is found by adding each team's offense and each opponent's defense.  Half the home field advantage coefficient is added to the home team and the other half is subtracted from the away team.  Both of these coefficients are then used as the exponent in the exponential function to find the means.  It will also calculate the probability of each of the three outcomes (home win, away, win, and draw.)

```{r}
max_goals = 6         # Specify maximum number of desired goals in joint distribution
ht = "Pittsburgh"         # Specify home team
at = "Nashville"          # Specify away team

exp_score = matrix(nrow = max_goals + 1, ncol = max_goals + 1, byrow = TRUE)  # Create empty matrix to hold joint distribution

# Find indices in coefficients corresponding to home and away teams
for(i in 1:length(teams)) {
  if(teams[i] == ht) { ht_index = i }
  if(teams[i] == at) { at_index = i }
}

# Calculate ratings for home and away teams 
home_score = exp(coefficients[1] + coefficients[2] + coefficients[2 + ht_index] - coefficients[2 + length(teams) + at_index])
away_score = exp(coefficients[1] + coefficients[2 + at_index] - coefficients[2 + length(teams) + ht_index])

# Initialize home win, away win, and draw probabilities
home_win_prob = 0
draw_prob = 0
away_win_prob = 0

for(i in 0:max_goals) {
  for(j in 0:max_goals) {
    # Calculate probability of home team scoring i goals and away team scoring j goals
    outcome_prob = ((home_score^i)*(exp(-home_score))/factorial(i)) * ((away_score^j)*(exp(-away_score))/factorial(j))
    exp_score[i+1,j+1] = outcome_prob
    
    # Add current probability to appropriate outcome
    if(i == j) {
      draw_prob = draw_prob + outcome_prob
    } else if (i < j) {
      away_win_prob = away_win_prob + outcome_prob
    } else {
      home_win_prob = home_win_prob + outcome_prob
    }
  }
}

# Creates scoreline distribution - cells inside table are percentages
scoreline_dist = round(exp_score * 100, 2)
colnames(scoreline_dist) = c(0:max_goals)
rownames(scoreline_dist) = c(0:max_goals)
scoreline_dist
```
# Elo Ratings

Calculate home field advantage - number of points added to the home team's Elo rating

```{r}
elos = read.table("nhl_initial_elos.csv", header = TRUE, sep = ",")
scores = read.table("nhl_data.csv", header = TRUE, sep = ",")

home_wins = 0
games = 0
first_game_index = 1400


# Iterate through games - first index can be changed to eliminate early seasons where scores are extreme
for(i in first_game_index:nrow(scores)) {  
  # Count number of games that do not end in ties
  if(scores$home_score[i] != scores$away_score[i] && scores$location[i] == 1) { games = games + 1 }
  
  # Count number of games where home team wins
  if(scores$home_score[i] > scores$away_score[i] && scores$location[i] == 1) { home_wins = home_wins + 1 }
}

home_win_prob = home_wins / games  # Calculate home win probability where outcome was not a tie
hfa = -400*log10(1/home_win_prob - 1)  # Calculate number of Elo points added to home team
```

This code chooses the optimal k-value by starting at the starting weight and incrementing up based on the specified step size until the number of iterations has been hit.  If the optimal k-value is one of the two endpoints, the code should be run again with different limits.  This is to ensure that the absolute minimum has been achieved.

If you already know the optimal k-value for a sport, this code does not need to be run.  The value can be hard coded in below when the Elo ratings are being calculated.

```{r}
elos = read.table("nhl_initial_elos.csv", header = TRUE, sep = ",")
scores = read.table("nhl_data.csv", header = TRUE, sep = ",")

starting_weight = 0  # Lower bound for weight ranges to be tested - generally set equal to 0
iterations = 100  # Number of k values to test
step_size = 0.1  # Amount to increment k by at each step
first_game_index = 9308

if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}


# Initialize data frame to store k values and corresponding error
errors = data.frame(matrix(ncol = 2, nrow = iterations))
colnames(errors) = c("weight", "error")
errors$weight = starting_weight + (1:iterations)*step_size
errors$error = NA

# Iterate through all potential k values that are being tested
for(counter in 1:iterations) {
    weight = starting_weight + counter*step_size  # Calculate k value for current iteration
    print(weight)
    error = 0  # Reset error for current iteration
    elos = read.table("nhl_initial_elos.csv", header = TRUE, sep = ",")  # Reset initial Elo ratings

    # Iterate through games - first index can be changed to eliminate early seasons in a league where early results tend to be extreme
    for(i in first_game_index:nrow(scores)) {  
        # Find indices corresponding to home and away teams for current game
        home_index = which(elos$team == scores$home_team[i])
        away_index = which(elos$team == scores$away_team[i])
        
        # Find home and away team Elo ratings
        home_elo = elos$rating[home_index]
        away_elo = elos$rating[away_index]
        
        # Calculate home team win probability
        win_prob = 1 / (10^((away_elo - (home_elo + hfa*scores$location[i]))/400) + 1)
        
        # Calculate actual margin of victory - must be positive
        score_diff = abs(scores$home_score[i] - scores$away_score[i])  
        
        # Determine home team result
        if(scores$home_score[i] > scores$away_score[i]) { 
          home_result = 1  # Home team wins
        } else if(scores$home_score[i] < scores$away_score[i]) { 
          home_result = 0  # Home team loses
        } else { 
          home_result = 0.5  # Tie
        }
        
        # Add squared error between home result and predicted probability of home team winning to SSE
        error = error + (home_result - win_prob)^2
        
        # Calculate amount each team's Elo rating is adjusted by
        home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)   # Log scaling function

        # Adjust Elo ratings - add point to winner and subtract points from loser
        elos$rating[home_index] = elos$rating[home_index] + home_elo_adjustment
        elos$rating[away_index] = elos$rating[away_index] - home_elo_adjustment

    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
        if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {
            for(j in 1:nrow(elos)) {
                if(scores$season[i] >= elos$inaugural_season[j]) {
                    elos$rating[j] = elos$rating[j] - (elos$rating[j] - 1500)/3
                }
            }
    
            existing_teams = elos[which(elos$inaugural_season <= (scores$season[i] + 1)),]
            expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
    
            for(j in 1:nrow(elos)) {
                if((scores$season[i] + 1) >= elos$inaugural_season[j]) {
                    elos$rating[j] = elos$rating[j] + expansion_adjustment
                }
            }
        }
    }
    errors$error[counter] = error  # Store error for current iteration
}

# Create plot of SSE against k-values
plot(errors$weight, errors$error, xlab = "k Value", ylab = "Sum of Squared Errors")

# Choose and print optimal weight based on value that had the lowest SSE
weight = errors$weight[which(errors$error == min(errors$error))]
print(weight)

plot(errors$weight, errors$error, pch = 19, xlab = "Weight", ylab = "Error")
```
After identifying the optimal k value (weight) and home field advantage, use those values to calculate the actual Elo ratings for each team in the league throughout its entire history.  This code also allows you to follow one team over a specified period of time by specifying the season and team below.

```{r}
# Read in initial team Elo ratings and history of games
elos = read.table("nhl_initial_elos.csv", header = TRUE, sep = ",")
scores = read.table("nhl_data.csv", header = TRUE, sep = ",")

#Running code to get scores at the end of the 2015 season
library(dplyr)

scores <- scores %>%
  filter(season < 2016)


# Input the optimal k factors (weight) and home field advantage
weight = 6.90
hfa = 51.11


# Identify if ties exist in the sport, which determines if an additional 0.5 needs to be added to the margin of victory
if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}


# Select team and season to follow over a specified period of time
team = "Pittsburgh Penguins"
first_season = 2016
last_season = 2016


# Create data frame to store information for team specified above
team_results = data.frame(matrix(ncol = 10, nrow = 0))
colnames(team_results) = c("opponent", "pregame_elo", "opponent_pregame_elo", "win_probability", "result", "team_score", "opponent_score", "elo_adjustment", "postgame_elo", "opponent_postgame_elo")

# Iterate through all games in the sport's history
for(i in 1:nrow(scores)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(elos$team == scores$home_team[i])
    away_index = which(elos$team == scores$away_team[i])
    
    
    # Find home and away team Elo ratings
    home_elo = elos$rating[home_index]
    away_elo = elos$rating[away_index]
    
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*scores$location[i]))/400) + 1)
    
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(scores$home_score[i] - scores$away_score[i])   
    
    
    # Determine home team result
    if(scores$home_score[i] > scores$away_score[i]) { 
        home_result = 1  # Home team wins
    } else if(scores$home_score[i] < scores$away_score[i]) { 
        home_result = 0  # Home team loses
    } else { 
        home_result = 0.5  # Tie
    }
  
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
    
    
    # Adjust Elo ratings - add point to winner and subtract points from loser
    elos$rating[home_index] = elos$rating[home_index] + home_elo_adjustment
    elos$rating[away_index] = elos$rating[away_index] - home_elo_adjustment

    
    # Add game information to team result data frame for each team game of the team specified above if team and season both match
    if(scores$season[i] >= first_season & scores$season[i] <= last_season & (scores$home_team[i] == team | scores$away_team[i] == team)) {
        if(scores$home_team[i] == team) {  # If specified team was at home
            team_results[nrow(team_results) + 1,] = c(scores$away_team[i], elos$rating[home_index] - home_elo_adjustment, elos$rating[away_index] + home_elo_adjustment, win_prob, home_result, scores$home_score[i], scores$away_score[i], home_elo_adjustment, elos$rating[home_index], elos$rating[away_index])
        } else {  # If specified team was away
            team_results[nrow(team_results) + 1,] = c(scores$home_team[i], elos$rating[away_index] + home_elo_adjustment, elos$rating[home_index] - home_elo_adjustment, 1-win_prob, 1-home_result, scores$away_score[i], scores$home_score[i], -1*home_elo_adjustment, elos$rating[away_index], elos$rating[home_index])
        }
    }
    
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {  # New season
        for(j in 1:nrow(elos)) {  # For each team
            if(scores$season[i] >= elos$inaugural_season[j]) {  # Check if team existed
                # Move each team's Elo rating back towards 1500 by 1/3 of the difference
                elos$rating[j] = elos$rating[j] - (elos$rating[j] - 1500)/3
            }
        }
        
        
        # Identify all teams that existed at beginning of following season
        existing_teams = elos[which(elos$inaugural_season <= (scores$season[i] + 1)),]
        
        
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
        
        
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(elos)) {  # For each team
            if((scores$season[i] + 1) >= elos$inaugural_season[j]) {  # Check if team existed
                elos$rating[j] = elos$rating[j] + expansion_adjustment  # Update ratings if so
            }
        }
    }
}


# Create plot of postgame Elo ratings
if (first_season == last_season) {
    team_name = paste(first_season, team)
} else {
    team_name = paste(paste(first_season, last_season, sep = "-"), team)
}

plot(team_results$postgame_elo, type = "l", xlab = team_name, ylab = "Elo Rating")
abline(h = 1500)
```


```{r}
elos_final$end_2015 <- elos$rating
```
Converting Elos to Start of 2016-2017 Season
```{r}
elos_final$start_2016 <- elos_final$end_2015 -(elos_final$end_2015 - 1500)/3
```
Renaming COlumns
```{r}
library(dplyr)
elos_final <- elos_final %>%
  rename(end_2016 = rating) %>%
  select(team, end_2015, start_2016, end_2016)
```

Visualizing the change
```{r}
elos_final$change = elos_final$end_2016 - elos_final$start_2016

rating_table <- elos_final[, c("team", "start_2016", "end_2016", "change")]

print(rating_table)
```
new table
```{r}
library(knitr)
kable(rating_table, caption = "NHL Elo Ratings 2016-2017")
```

# Simulating Seasons

Initialize teams and ratings

```{r}
# Specify season to simulate
simulated_season = 2016

# Read in list of all games
scores = read.table("nhl_data.csv", header = TRUE, sep = ",")

# Read in initial Elo rating, conference, and division
team_info = read.table("nhl_initial_elos.csv", header = TRUE, sep = ",")

# Obtain list of unique conference names and unique division names
conferences = na.omit(unique(team_info$conference))
divisions = na.omit(unique(team_info$division))

# Create list of games that occurred prior to season being simulated
pre_season = scores[which(scores$season < simulated_season & scores$season >= 1901),]

# Create list of regular season games for season being simulated
season_schedule = scores[which(scores$season == simulated_season & (scores$game_type == "r" | scores$game_type == "np")),]
```

Calculate initial Elo ratings

```{r}
# Input the optimal k factors (weight) and home field advantage
hfa = 51.11
weight = 6.9

# Identify if ties exist in the sport, which determines if an additional 0.5 needs to be added to the margin of victory
if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}

# Iterate through all games in the sport's history up to season being simulated
for(i in 1:nrow(pre_season)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(team_info$team == pre_season$home_team[i])
    away_index = which(team_info$team == pre_season$away_team[i])
    
    # Find home and away team Elo ratings
    home_elo = team_info$rating[home_index]
    away_elo = team_info$rating[away_index]
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*pre_season$location[i]))/400) + 1)
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(pre_season$home_score[i] - pre_season$away_score[i])
    
    # Determine home team result
    if(pre_season$home_score[i] > pre_season$away_score[i]) { 
    home_result = 1  # Home team wins
    } else if(pre_season$home_score[i] < pre_season$away_score[i]) { 
    home_result = 0  # Home team loses
    } else { 
    home_result = 0.5  # Tie
    }
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)

  
    # Adjust Elo ratings - add point to winner and subtract points from loser
    team_info$rating[home_index] = team_info$rating[home_index] + home_elo_adjustment
    team_info$rating[away_index] = team_info$rating[away_index] - home_elo_adjustment
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {
        for(j in 1:nrow(team_info)) {
            if(scores$season[i] >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] - (team_info$rating[j] - 1500)/3
                }
        }
    
        # Identify all teams that existed at beginning of following season
        existing_teams = team_info[which(team_info$inaugural_season <= (scores$season[i] + 1)),]
    
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
    
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(team_info)) {
            if((scores$season[i] + 1) >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] + expansion_adjustment
            }
        }
    }
}
```

Simulate season

```{r}
# Set seed for replication purposes
set.seed(31)

# Determine number of times to simulate the season
iterations = 10000

# Create data frame to hold Elo ratings, actual win totals, and simulation results
results = data.frame(matrix(0, ncol = 6, nrow = nrow(team_info)))
colnames(results) = c("team", "starting_elo", "ending_elo", "actual_wins", "average_wins", "division_titles")
results$team = team_info$team

# Create data frame to hold number of wins by each team in each iteration
win_totals = data.frame(matrix(0, ncol = nrow(team_info), nrow = iterations))
colnames(win_totals) = team_info$team

# Simulate the season the given number of times
for(i in 1:iterations) {
    if(i %% 1000 == 0) {print(i)}
    season_stats = team_info[,which(colnames(team_info) != "inaugural_season")]
    season_stats$wins = 0
    season_stats$rand = runif(nrow(team_info))
    
    # Simulate each game in current season
    for(j in 1:nrow(season_schedule)) {
        # Find indices corresponding to home and away teams for current game
        home_index = which(season_stats$team == season_schedule$home_team[j])
        away_index = which(season_stats$team == season_schedule$away_team[j])
        
        # Find home and away team Elo ratings
        home_elo = season_stats$rating[home_index]
        away_elo = season_stats$rating[away_index]
        
        # Calculate home team win probability
        win_prob = 1 / (10^((away_elo - (home_elo + hfa*season_schedule$location[j]))/400) + 1)
        u = runif(1)  # Generate a random number used to determine the winner of the game
        
        # Determine which team wins the simulated game and increment their win total by 1
        if(u < win_prob) {
            season_stats$wins[home_index] = season_stats$wins[home_index] + 1
        } else {
            season_stats$wins[away_index] = season_stats$wins[away_index] + 1
        }
    
        # Calculate actual margin of victory - must be positive
        score_diff = abs(season_schedule$home_score[j] - season_schedule$away_score[j])
        
        # Determine home team result
        if(season_schedule$home_score[j] > season_schedule$away_score[j]) { 
          home_result = 1  # Home team wins
        } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) { 
          home_result = 0  # Home team loses
        } else { 
          home_result = 0.5  # Tie
        }
        
        # Calculate amount each team's Elo rating is adjusted by
        home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
        
        # Adjust Elo ratings after game has been simulated to get team's new strength
        season_stats$rating[home_index] = season_stats$rating[home_index] + home_elo_adjustment
        season_stats$rating[away_index] = season_stats$rating[away_index] - home_elo_adjustment
    
        # Track season wins
        if(i == 1) {
            if(season_schedule$home_score[j] > season_schedule$away_score[j]) {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 1
            } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) {
                results$actual_wins[away_index] = results$actual_wins[away_index] + 1
            } else {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 0.5
                results$actual_wins[away_index] = results$actual_wins[away_index] + 0.5
            }
        }
    }

    # Define data frame that contains division winners
    division_winners = data.frame(matrix(ncol = 6, nrow = 0))
    colnames(division_winners) = c("team", "conference", "division", "rating", "wins", "rand")
  
    # For each division
    for(div in divisions) {
        div_standings = season_stats[which(season_stats$division == div),]  # Identify all teams in current division
        div_standings = div_standings[order(-div_standings$wins, -div_standings$rand),]  # Sort division by wins and random number
        division_winners = rbind(division_winners, div_standings[1,])  # Add division winner to 'division_winners' data frame
    }
  
    # Save end of season win totals (from column 5 of season_stats) for each iteration in the win_totals data frame
    win_totals[i,] = t(season_stats[5])
    
    # Add number of wins for each team during this iteration to sum
    results$average_wins = results$average_wins + season_stats$wins
  
    # Increment the number of division titles for each division winner by 1
    for(team in division_winners$team) {
        index = which(season_stats$team == team)  # Index of division winner
        results$division_titles[index] = results$division_titles[index] + 1  # Increment division titles
    }
}
 
# Calculate average number of wins across all iterations
results$starting_elo = team_info$rating
results$ending_elo = season_stats$rating
results$average_wins = results$average_wins/iterations
results$division_titles = results$division_titles/iterations
```
Visualizing Table
```{r}
results$residual = results$actual_wins - results$average_wins

library(dplyr)

results_plots <- results %>%
  filter(residual != 0)

hist(results_plots$residual, main = "Histogram of Residuals", xlab = "Actual Games Won - Simulated Games Won", breaks = 7)
plot(results_plots$actual_wins, results_plots$average_wins, pch = 19, xlab = "Actual Wins", ylab = "Simulated Wins", main = "Actual vs Simulated Wins 2016-17")
cor(results_plots$actual_wins, results_plots$average_wins)
```
Edmonton Histogram of Simulated Wins
```{r}
hist(win_totals$`Edmonton Oilers`, main = "Histogram of Edmonton Oilers Simulated Win Totals", xlab = "Win Totals")
```
```{r}
library(dplyr)

s_2016_good <- s_2016 %>%
  select(Team, wpct, pyth_wpct, error1) %>%
  mutate(actual_points = wpct * 164,
         pyth_points = pyth_wpct * 164)
```
 Stats for edmonton
```{r}
mean(win_totals$`Edmonton Oilers`)
sd(win_totals$`Edmonton Oilers`)
```
```{r}
(47-38.5532) / 4.41
```
 Avs hist
 hist(win_totals$`Edmonton Oilers`, main = "Histogram of Edmonton Oilers Simulated Win Totals", xlab = "Win Totals")
```{r}
hist(win_totals$`Colorado Avalanche`, main = "Histogram of Colorado Avalanche Simulated Win Totals", xlab = "Win Totals")
```
 summary avs
```{r}
sd(win_totals$`Colorado Avalanche`)
mean(win_totals$`Colorado Avalanche`)
```
```{r}
(22-32.13) / 4.31
```
 



